\Large \bf Cyclone: High Availability for Persistent Key Value Stores

### 2017-08-10 paper/paper.tex
Spellcheck.

These key value stores however tend to ignore an important component: the write
ahead log. A machine or storage failure leading to total loss or temporary
unavailability of data is unacceptable in services where high availabilty and
revenue are interconnected. Key value stores therefore usually incorporate
support for a write ahead log that if replicated and kept durable for every
appended update provides the necessary high availability. Unfortunately, the
write ahead log is a performance achilles heel for these systems, eclipsing much
of the work on improving the performance of the LSM component. To illustrate the
impact of the write ahead log, consider Figure~\ref{fig:problem}. The line
marked 'Rocksdb' shows the performance of Rocksdb without the write ahead
log. The performance when persisting the write ahead log without replicating it
is shown as the line marked 'rocksdb/WAL'. The line marked 'rocksdb/3 way rep.'
is for simply replicating the log three ways without persisting it, using
RAFT~\cite{raft} running over TCP/IP. Either persisting every update to the log
or replicating it using TCP/IP causes performance to drop by an order of
magnitude (note the log scale on the x-axis). It is therefore no surprise that
deployments of Rocksdb often turn off the write ahead log~\cite{samza},
depending on upper layers to provide availability by co-ordinating replicas. On
the other side of the spectrum, key value store research prototypes such as
FloDB~\cite{flodb} turn off the write ahead log to be able to showcase benefits
of sophisticated extensions to LSM data structures.

@@

The rest of the paper is organized as follows. We describe Cyclone's system
architecture in Section~\ref{sec:sysarch}. We describe the two level log
structure in Section~\ref{sec:storage}. The set of optimizations to the network
stack is decribed in Section~\ref{sec:network}. We show how Cyclone can be
scaled to make use of available parallelism on the system in
Section~\ref{sec:parallelism}. A detailed evaluation of Cyclone's replication
performance is provided in the context of RocksDB - a popular persistent key
value store - in Section~\ref{sec:evaluation}. We then discuss related work
before concluding.

@@

\section{Storage}
\label{sec:storage}
Storage devices such as flash-based SSDs export a block IO interface. Appending
a small update to the write-ahead log requires a synchronous access to the
SSD. Even with write coalescing using a write cache on the SSD, the round trip
time to a current generation NVMe SSD such as the one we use in this
paper~\cite{ssd_spec}, is of the order of 20 us, limiting the throughput to
under 50K ops/sec and resulting in a high baseline latency at even low load (as
shown in Figure~\ref{fig:problem}). This problem leads many system designers to
underprovision key-value store shards ensuring that it operates at moderate to
high load, using group-commit to batch additions to the log. They therefore pay
a price in latency to ensure the bandwidth to storage is fully utilized.

@@

DPDK does not by itself provide a TCP stack. But this is not a problem, since
RAFT (and indeed most consensus protocols) tolerate network losses and
reordering by design due to the need to support asynchronous communication. In
addition, most datacenter networks rarely drop or reorder packets and provide
full bisection bandwith between servers that might serve as Cyclone replicas. We
therefore jettisoned TCP and chose to send raw IP packets encapsulated in
ethernet frames. Cyclone takes complete control of an ethernet interface,
receiving all packets directed to it, while the IP and ethernet addresses
provide sufficient information to route the packet if necessary through multiple
switches. We currently follow this communication model both for server to server
communication between replicas as well as client to server communication for
requests and responses. Although a detailed evaluation is made later, switching
from the kernel TCP/IP stack to DPDK reduces the latency between machines in our
testbed from {\tt 18 us} to as low as {\tt 5 us}, providing a significant boost
to performance.

@@

The first question we deal with is -- how do Cyclone clients decide which
physical log to send a request to? A key constraint here is that all replicas
should agree on the seqeunce of updates to a key, as otherwise they would
diverge in terms of state. Allowing updates to the same key to go to different
physical logs would lead to a divergence as there is no guarantee about the
order in which the independent RAFT instances would apply the update. Therefore
all updates to a key must go to the same physcial log. A similar constraint
applies to reads in order to ensure that all committed writes are seen by
readers. Allowing reads to a key to go to a different RAFT instance than the
write means that it could be served on a machine where the write is yet to be
applied. Therefore, all reads and writes to a key must go to the same physical
log. We achieve this by hashing the key to select the physical log.

@@

The first key problem we need to deal with is that the leaders for the different
RAFT instances to which the keys map can be located on different machines. To
avoid having a client co-ordinate different machines for a single operation, we
constrain the RAFT leader election algorithm for the different RAFT instances to
converge to the same machine. We do this by triggering re-elections for a RAFT
instance other than that of the first physical log until it is on the same
machine as the \emph{first} physical log. This process is therefore resilient to
failed machines. RAFT requires that the leader have the most up to date log. To
ensure that the process converges we therefore ensure that the currently elected
leader brings a majority quroum up to date before triggering a relection. We
also do not accept any requests from clients until all leaders are
co-located. We assume that no network partitions occur \emph{within a machine}
i.e. some RAFT instances are able to send and receive packets, while others are
not, a possiblity we discount due to our single process design.

@@

A failure causes a new leader to be elected for the affected physical log,
moving forward the term. If the participants of a ganged operatation do not
detect this case, they could be left waiting forever. We detect failure on any
RAFT instance by having each instance publish its current term and continuously
comparing the view in the ganged operation to it. If any RAFT instance has moved
past that term, the ganged operation is then terminated.

@@

The assumption in Figure~\ref{fig:ganged_ops} is that each ganged operation is
mapped to a unique barrier. We achieve this by using a fixed piece of memory
owned by the co-ordinator to hold the barrier and write the nonce to it in order
to indicate that the barrer is active for the corresponding ganged
operation. Participants watch for the nonce to know when to execute the
ganged operation barrier in Figure~\ref{fig:ganged_ops}, while also monitoring
the leader's published view to detect the case where the ganged operation fails
to replicate on the co-ordinator's physical log.

@@

Ganged operations possibly consititute the most complex part of Cyclone but the
code weighs in at well under a couple of hundred lines. We believe that
this additional complexity is still small compared to distributed transactions
using two-phase commit - especially when taking into account failure recovery -
as used in systems such as FARM~\cite{farm}.

@@

One can therefore conclude that using a small amount of directly attached NVM is
effective at hiding the latency of a background block storage device from the
criticial path of replication. Further, for small updates the
bottleneck is the capability of the software to run the consensus protocol, but
the bottleneck shifts to the secondary storage device for larger request sizes.

We now evaluate ganged operations. The primary purpose of
ganged operations is to avoid the need for distributed transactions to
manipulate what is a single shared memory image and therefore we were most
concerned about unloaded latency given the complexity of synchronizing different
replication quorums as well executing our rendezvous protocol on multiple
cores. We therefore setup an experiment where a single client -- reflecting the
unloaded case -- made ganged requests to the replicas. We varied the number of
physical logs participating in the request from one to the full complement of 8.
Figure~\ref{fig:ganged} shows the results. Unloaded
latency increases slowly as we increase the number of active physical logs - to
around 32 us from the baseline of 21 us. The reason for this increase is that
the DPDK userspace NIC driver pipelines request processing for communication
between the CPU core and the NIC. Simultaneous replicaton on multiple physical
logs represents a worst case for this arrangement as all eight cores try to
send replication messages to their replicas, adding about a microsecond of
serial latency each, to dispatch requests to the NIC. Regardless, the experiment
underlines the value of our ganged operation design over a distributed
transaction that would need 8 network hops for the two phase commit (and
replication) using an external co-ordinator, a minimum of 80us.

@@

The performance of Rocksdb with Cyclone for the small update workload is shown
in Figure~\ref{fig:rocksdb} - essentially presenting the solution to the problem
demonstrated in Figure~\ref{fig:problem}.  We consider four different
setups. The line labeled 'rocksdb' is the key value store running with no
logging whatsover - a system crash would lead to data loss. The line labeled
'rocksdb/wal' is for Rocksdb running with its write ahead logging turned on. The
large gap between these two is the overhead of the existing Rocksdb WAL
solution. The line labeled 'rocksdb/Cyclone 1 way' is a two level Cyclone log
but without any replication. The line almost exactly tracks the performance of
Rockdb. Cyclone is able to provide a write ahead log with no overhead to
Rocksdb. The line labeled 'rocksdb/Cyclone 3 way' is with 3-way replication
turned on. Other than a 20us delta due to the extra network round trip, the line
almost exactly tracks Rocksdb performance with no logging. Cyclone therefore
provides high availability to Rocksdb at a fraction of the cost of its existing
single machine write ahead log. We also repeat the experiment for the larger
update size in Figure~\ref{fig:rocksdb_256}. The conclusions are identical:
Cyclone solves Rocksdb's write ahead logging problem.

@@

Consensus in a box~\cite{consensus_box} implements the Zookeeper atomic
broadcast protocol together with a NIC on an FPGA. That work ignored durability,
focusing purely on network performance for replication. Latency is excellent due
to the fact that FPGA cuts out the path from CPU caches to the NIC. The authors
report latency as low as 3us for a round trip from leader to follower replica,
compared to the 6us we observe with DPDK. Notably however they reported 7us when
using TCP/IP from the FPGA rather than their own connection oriented network
protocol. On the other hand their peak replication throughput is 4M
replications/sec on 66 Gbps of aggregate network bandwith for small messages,
comparable to the 6M replications/sec mark obtained by Cyclone on an aggregate
40 Gbps of network bandwidth when \emph{also persisting the log on the SSD}. We
believe the general applicability afforded by our software only approach
together with the fact that we provide persistence by design makes it a
compelling alternative to an FPGA based solution, even given the larger
latency. Also, these results demonstrate that a general purpose CPU core can be
as efficient as an FPGA in running a consensus protocol using the techniques
described in this paper.

@@

{CRANE}~\cite{crane} is a system for replicating mutithreaded programs using
Paxos. It uses deterministic multi-threading to ensure replicas converge to the
same state as opposed to semantic equivalence in Cyclone using API
commutativity.  CRANE reports around a 2X slowdown for MySQL, likely rendering
it unsuitable as a replacement for the write ahead log in key value stores.

@@

Database practitioners have considered the utility of byte adressable
non-volatile memory in improving logging for
databases~\cite{nvram_log}. However, that work proposes placing the entire log
in NVM, presuming sufficient availability of such memory. It also does not
consider replication.

>>>>

These key value stores however tend to ignore an important component: the write
ahead log. A machine or storage failure leading to total loss or temporary
unavailability of data is unacceptable in services where high availability and
revenue are interconnected. Key value stores therefore usually incorporate
support for a write ahead log that if replicated and kept durable for every
appended update provides the necessary high availability. Unfortunately, the
write ahead log is a performance achilles heel for these systems, eclipsing much
of the work on improving the performance of the LSM component. To illustrate the
impact of the write ahead log, consider Figure~\ref{fig:problem}. The line
marked 'Rocksdb' shows the performance of Rocksdb without the write ahead
log. The performance when persisting the write ahead log without replicating it
is shown as the line marked 'rocksdb/WAL'. The line marked 'rocksdb/3 way rep.'
is for simply replicating the log three ways without persisting it, using
RAFT~\cite{raft} running over TCP/IP. Either persisting every update to the log
or replicating it using TCP/IP causes performance to drop by an order of
magnitude (note the log scale on the x-axis). It is therefore no surprise that
deployments of Rocksdb often turn off the write ahead log~\cite{samza},
depending on upper layers to provide availability by co-ordinating replicas. On
the other side of the spectrum, key value store research prototypes such as
FloDB~\cite{flodb} turn off the write ahead log to be able to showcase benefits
of sophisticated extensions to LSM data structures.

@@

The rest of the paper is organized as follows. We describe Cyclone's system
architecture in Section~\ref{sec:sysarch}. We describe the two level log
structure in Section~\ref{sec:storage}. The set of optimizations to the network
stack is described in Section~\ref{sec:network}. We show how Cyclone can be
scaled to make use of available parallelism on the system in
Section~\ref{sec:parallelism}. A detailed evaluation of Cyclone's replication
performance is provided in the context of RocksDB - a popular persistent key
value store - in Section~\ref{sec:evaluation}. We then discuss related work
before concluding.

@@

\section{Storage}
\label{sec:storage}
Storage devices such as flash-based SSDs export a block IO interface. Appending
a small update to the write-ahead log requires a synchronous access to the
SSD. Even with write coalescing using a write cache on the SSD, the round trip
time to a current generation NVMe SSD such as the one we use in this
paper~\cite{ssd_spec}, is of the order of 20 us, limiting the throughput to
under 50K ops/sec and resulting in a high baseline latency at even low load (as
shown in Figure~\ref{fig:problem}). This problem leads many system designers to
under-provision key-value store shards ensuring that it operates at moderate to
high load, using group-commit to batch additions to the log. They therefore pay
a price in latency to ensure the bandwidth to storage is fully utilized.

@@

DPDK does not by itself provide a TCP stack. But this is not a problem, since
RAFT (and indeed most consensus protocols) tolerate network losses and
reordering by design due to the need to support asynchronous communication. In
addition, most datacenter networks rarely drop or reorder packets and provide
full bisection bandwidth between servers that might serve as Cyclone replicas. We
therefore jettisoned TCP and chose to send raw IP packets encapsulated in
ethernet frames. Cyclone takes complete control of an ethernet interface,
receiving all packets directed to it, while the IP and ethernet addresses
provide sufficient information to route the packet if necessary through multiple
switches. We currently follow this communication model both for server to server
communication between replicas as well as client to server communication for
requests and responses. Although a detailed evaluation is made later, switching
from the kernel TCP/IP stack to DPDK reduces the latency between machines in our
testbed from {\tt 18 us} to as low as {\tt 5 us}, providing a significant boost
to performance.

@@

The first question we deal with is -- how do Cyclone clients decide which
physical log to send a request to? A key constraint here is that all replicas
should agree on the sequence of updates to a key, as otherwise they would
diverge in terms of state. Allowing updates to the same key to go to different
physical logs would lead to a divergence as there is no guarantee about the
order in which the independent RAFT instances would apply the update. Therefore
all updates to a key must go to the same physical log. A similar constraint
applies to reads in order to ensure that all committed writes are seen by
readers. Allowing reads to a key to go to a different RAFT instance than the
write means that it could be served on a machine where the write is yet to be
applied. Therefore, all reads and writes to a key must go to the same physical
log. We achieve this by hashing the key to select the physical log.

@@

The first key problem we need to deal with is that the leaders for the different
RAFT instances to which the keys map can be located on different machines. To
avoid having a client co-ordinate different machines for a single operation, we
constrain the RAFT leader election algorithm for the different RAFT instances to
converge to the same machine. We do this by triggering re-elections for a RAFT
instance other than that of the first physical log until it is on the same
machine as the \emph{first} physical log. This process is therefore resilient to
failed machines. RAFT requires that the leader have the most up to date log. To
ensure that the process converges we therefore ensure that the currently elected
leader brings a majority quorum up to date before triggering a reelection. We
also do not accept any requests from clients until all leaders are
co-located. We assume that no network partitions occur \emph{within a machine}
i.e. some RAFT instances are able to send and receive packets, while others are
not, a possibility we discount due to our single process design.

@@

A failure causes a new leader to be elected for the affected physical log,
moving forward the term. If the participants of a ganged operation do not
detect this case, they could be left waiting forever. We detect failure on any
RAFT instance by having each instance publish its current term and continuously
comparing the view in the ganged operation to it. If any RAFT instance has moved
past that term, the ganged operation is then terminated.

@@

The assumption in Figure~\ref{fig:ganged_ops} is that each ganged operation is
mapped to a unique barrier. We achieve this by using a fixed piece of memory
owned by the co-ordinator to hold the barrier and write the nonce to it in order
to indicate that the barrier is active for the corresponding ganged
operation. Participants watch for the nonce to know when to execute the
ganged operation barrier in Figure~\ref{fig:ganged_ops}, while also monitoring
the leader's published view to detect the case where the ganged operation fails
to replicate on the co-ordinator's physical log.

@@

Ganged operations possibly constitute the most complex part of Cyclone but the
code weighs in at well under a couple of hundred lines. We believe that
this additional complexity is still small compared to distributed transactions
using two-phase commit - especially when taking into account failure recovery -
as used in systems such as FARM~\cite{farm}.

@@

One can therefore conclude that using a small amount of directly attached NVM is
effective at hiding the latency of a background block storage device from the
critical path of replication. Further, for small updates the
bottleneck is the capability of the software to run the consensus protocol, but
the bottleneck shifts to the secondary storage device for larger request sizes.

We now evaluate ganged operations. The primary purpose of
ganged operations is to avoid the need for distributed transactions to
manipulate what is a single shared memory image and therefore we were most
concerned about unloaded latency given the complexity of synchronizing different
replication quorums as well executing our rendezvous protocol on multiple
cores. We therefore setup an experiment where a single client -- reflecting the
unloaded case -- made ganged requests to the replicas. We varied the number of
physical logs participating in the request from one to the full complement of 8.
Figure~\ref{fig:ganged} shows the results. Unloaded
latency increases slowly as we increase the number of active physical logs - to
around 32 us from the baseline of 21 us. The reason for this increase is that
the DPDK userspace NIC driver pipelines request processing for communication
between the CPU core and the NIC. Simultaneous replication on multiple physical
logs represents a worst case for this arrangement as all eight cores try to
send replication messages to their replicas, adding about a microsecond of
serial latency each, to dispatch requests to the NIC. Regardless, the experiment
underlines the value of our ganged operation design over a distributed
transaction that would need 8 network hops for the two phase commit (and
replication) using an external co-ordinator, a minimum of 80us.

@@

The performance of Rocksdb with Cyclone for the small update workload is shown
in Figure~\ref{fig:rocksdb} - essentially presenting the solution to the problem
demonstrated in Figure~\ref{fig:problem}.  We consider four different
setups. The line labeled 'rocksdb' is the key value store running with no
logging whatsoever - a system crash would lead to data loss. The line labeled
'rocksdb/wal' is for Rocksdb running with its write ahead logging turned on. The
large gap between these two is the overhead of the existing Rocksdb WAL
solution. The line labeled 'rocksdb/Cyclone 1 way' is a two level Cyclone log
but without any replication. The line almost exactly tracks the performance of
Rocksdb. Cyclone is able to provide a write ahead log with no overhead to
Rocksdb. The line labeled 'rocksdb/Cyclone 3 way' is with 3-way replication
turned on. Other than a 20us delta due to the extra network round trip, the line
almost exactly tracks Rocksdb performance with no logging. Cyclone therefore
provides high availability to Rocksdb at a fraction of the cost of its existing
single machine write ahead log. We also repeat the experiment for the larger
update size in Figure~\ref{fig:rocksdb_256}. The conclusions are identical:
Cyclone solves Rocksdb's write ahead logging problem.

@@

Consensus in a box~\cite{consensus_box} implements the Zookeeper atomic
broadcast protocol together with a NIC on an FPGA. That work ignored durability,
focusing purely on network performance for replication. Latency is excellent due
to the fact that FPGA cuts out the path from CPU caches to the NIC. The authors
report latency as low as 3us for a round trip from leader to follower replica,
compared to the 6us we observe with DPDK. Notably however they reported 7us when
using TCP/IP from the FPGA rather than their own connection oriented network
protocol. On the other hand their peak replication throughput is 4M
replications/sec on 66 Gbps of aggregate network bandwidth for small messages,
comparable to the 6M replications/sec mark obtained by Cyclone on an aggregate
40 Gbps of network bandwidth when \emph{also persisting the log on the SSD}. We
believe the general applicability afforded by our software only approach
together with the fact that we provide persistence by design makes it a
compelling alternative to an FPGA based solution, even given the larger
latency. Also, these results demonstrate that a general purpose CPU core can be
as efficient as an FPGA in running a consensus protocol using the techniques
described in this paper.

@@

{CRANE}~\cite{crane} is a system for replicating multithreaded programs using
Paxos. It uses deterministic multi-threading to ensure replicas converge to the
same state as opposed to semantic equivalence in Cyclone using API
commutativity.  CRANE reports around a 2X slowdown for MySQL, likely rendering
it unsuitable as a replacement for the write ahead log in key value stores.

@@

Database practitioners have considered the utility of byte addressable
non-volatile memory in improving logging for
databases~\cite{nvram_log}. However, that work proposes placing the entire log
in NVM, presuming sufficient availability of such memory. It also does not
consider replication.
